---
title: 'Social links and upward mobility: analyzing the patrician social network of
  the Republic of Venice'
author: "JJ Merelo"
date: '2022-08-23'
output: pdf_document
abstract:
  The Venetian republic was one of the most successful transmodern states, surviving for a millenium through innovation, commercial cunning, exploitation of colonies and legal stability. Part of the success might be due to its government structure, a republic ruled by a doge chosen among a set of Venetian patrician families. In this paper we analyze the structure of the social network they formed through marriage, and how government was monopolyzed by a relatively small set of families, the one that became patrician first.
bibliography: ../patrician.bib
---

```{r setup, include=FALSE}
library(igraph)
data <- read.csv("../resources/family-pairs.csv",header=FALSE)
venice.sn <- graph.data.frame(data.frame(data$V1,data$V2),directed=F)
```


## Introduction

During one thousand years, the Venetian republic managed to raise from a small fiefdom under suzerainty of the Byzantine empire to the biggest empire in the Mediterranean, and then maintain its existence until the *Napoleonic* storm reconfigured the map of Europe for ever [@horodowich2013brief].

One of the main factors that influenced in that stability was the way it chose its presidents, or *doges*, as well as the constraints put on what they could effectively do, as well as a clear separation of powers that created a plethora of institutions with more or less clearly defined functions [@institutions]. The mercantile orientation of these institutions certainly helped, since it created a safe and secure environment where people from all over the world could do business. But it was extremely rare for a doge to be deposed due to corruption, or ousted by a coup d'ètat, as it was extremely common all over the Italian peninsula and elsewhere.

However, Venice was a far cry from a democracy. Nobles (*patriziato*) and the people (*popolani*) were totally separate, with the patricians being the only one with a possibility of occupying any job going from simple administrative jobs, to the highest offices such as the Maggior Consiglio (equivalent to a Privy Council), the Senate or the *Consiglio dei dieci*, that worked as higher instance tribunal as well as home office. This effectively implied that a certain number of families ruled the Republic, from its inception to the end. This rule was, also, increasingly authoritarian, with a tight grip on power, that was, however, distributed among the different institutions.

It is, then, difficult to understand how what caused upheaval and instability in most Medieval and post-Medieval states instead was not a factor leading to the demise of the institution of the Doges. This paper tries to contribute to understand why, by studying the social network formed by these patrician families, and how it is related to the number of doges these families contributed as governors of the republic.

## Dataset generation

It is theoretically possible, but out of reach for people that are not scholars, to access the whole marriage registry in the city of Venice. It is relatively simple, however, to find the list of all doges and all the women they were married to. Initially, I thought it was possible to use a simple Wikidata query to do so, but unfortunately those Wikidata predicates have not been populated yet. So the dataset had to be built pretty much by hand.

The data set has been created specifically from this line of research. The list of all doges, and the list of all *dogaresse*, or doge's wives, has been extracted from Wikipedia pages, the list of doges from the Spanish page (since it was already formatted in a list) and the list of dogaresse from the Italian Wikipedia. These lists included dates when they were dogaresse; they were matched to their husbands by these dates. Couples were then double-checked, when some possible doubt (due to dates) was found, using primary sources such as [@cicogna1863storia]. In some cases, dates had to be fixed in the original list, as is shown [in this diff](https://it.wikipedia.org/w/index.php?title=Dogaressa&type=revision&diff=128805136&oldid=127173404). The fact that Venetian wives did keep their family names, or at least were recorded that way, helps to identify them as members of a specific family.

```{r raw, include=FALSE}
all.data <- read.csv("../resources/Venice-marriage-dogi.csv",header=FALSE)
married.data <- all.data[ all.data$V2 != "",]
```

All in all, there were `r nrow(all.data)` doges through all history. Out of all of these, `r nrow(all.data) - nrow(married.data)` did not marry, or their marriage has not been registered. Out of these couples, there were a few doges that married twice; most married only once. Additional filtering and processing was done:

* Couples marrying into the same family were eliminated. Patrician families were quite extensive, and sometimes had different branches. it was not common, but it happened, that some doge was married to a dogaressa belonging to another branch of the family. For instance, Pietro III Candiano as well as Pietro IV Candiano married to women of the Candiano family; Domenico Selvo married in 1075 to Theodora Selvo; Ordelafo Falier married to Matelda Falier, and right after that one, Domenico Michele married to Alicia Michele.
* The patrician family was identified by following a regular expression on the original data. The surname of the doge was registered as the last name before the date of rule, the dogaresse was the last expression in the string. However, the extracted surname did not always correspond to a patrician family. In some cases, mainly in the initial stages of the republic, only first names were known. In some other case, mainly in the case of dogaresse, only their country or land of origin was noted: for instance, Valdrada di Sicilia or Loicia da Prata were married to doges in the late XIII century.
* Family names were canonicalized. Many patrician family names have two forms: the one in Veneto, the language spoken in Venice, and the one in Italian. For instance, Cornaro and Corner, or Michele and Michel. In a case, "Mastropiero", it was also known as "Malipiero"; Gradenigo could also be written as Gradenico. These were normalized to a single form, in Veneto, which is the one found in most primary sources.

```{r sets, include=FALSE}
unique.doges.pairs <- unique(data$V1)
unique.dogaresse <- unique(data$V2)
both <- intersect(unique.doges.pairs,unique.dogaresse)
any.d <- union( unique.doges.pairs, unique.dogaresse)
```

After these filters, there were only `r nrow(data)` left. Out of these, there were only `r nrow(unique.doges.pairs)` different families that became "Serenest princes" or doges, `r nrow(unique.dogaresse)` that were *dogaresse*. Some families had members in both roles: `r length(both)` of them, a very small percentage of the total families; on the other hand, a total of `r length(any.d)` participated in any of them.

```{r doges, include=FALSE}
doges <- read.csv("../resources/doge-surnames.csv")
non.married <- setdiff( doges$Family, unique.doges.pairs )
```

Out of all the families that contributed to the top most office in the republic, there were `r length(non.married)` that did not marry, or their marriages were not registered. Some relatively important families, like Bembo, Erizzo or Michele, appear in this list. 
Five o them, including Bembo, do appear as dogaresse. This means that out of the 65 families that became doges, just 26 are not linked by marriage to ther families, at least with this small sample of links that is the list of married doges. Most of these non-married families, except five of them, Partecipazio, Candiano, Ziani, Galbaio and Teodato, had a single doge to their account. These are special cases: Partecipazio were the first doges, when office was still kind of inherited (the doge nominated a "consul", which succeeded him). So they *had* to be married; same goes for Candiano, except these took office a bit later. Galbaio and Teodato were not even properly doges, but more dukes or military leaders that took office in the first millennium. Ziani was in fact married, except that he was married to one such Cecilia, with no family recorded. So in fact families with more than one doge in them did include at least one marriage link with other patrician families; as a matter of fact, the Bembo family, for instance, included only one doge, Giovanni Bembo, but two dogaresse, Cornelia and Felicità.

We can draw two conclusions from this: being married was the prevalent civil state of the doges, and the marriage registry is a good enough tool to research the social network of the patrician families of which they sprung.

## Conclusions

The complicated way in which the doges were elected, and how individuals were chosen to enter a career as officers of the Republic, guaranteed that extremely wide, and variable, alliances had to be put in play. These alliances were solidified through marriages, which integrated politically as well as commercially the families of the two (romantic and commercial) partners.

In this paper we have drawn two conclusions from the study of the social network. One, that *old* (*vecchie*) families had a firm grasp on power, through alliances that were kept in places for centuries. Two, that the position in the social network was highly correlated with their position in an hypothetical ranking of the positions of powers held by that family through the centuries. More than individual contributions (which were anyway important to be able to marry into one of the best families), how links between families had been forged and kept were the ones that made families raise to the top job in the Republic over and over again over the centuries.

This, again, shows that quantitative studies over social networks are an excellent tool for understanding political history of different countries, and contribute to the explanation of the stability of the Republic's government: by raising to power only persons in families that were able to forge the widest alliances, *popular* will was satisfied (as a matter of fact, *patrician* will, since they were the ones that effectively had the power to vote), and this was one of the factors that contributed to the stability way beyond the zenith of the Venetian empire.



## References